Scattering from compact objects: \\ Regge poles and the Complex Angular Momentum method

### 2019-12-06 rainbowCAM.tex
reference added for Decanini and initial conditions for Dn

%\usepackage[margin=1in]{geometry}
\usepackage{float}
%\usepackage{subfig}
\usepackage{comment}
\usepackage{array}
%\usepackage{varioref}
%\usepackage[hidelinks]{hyperref}
%\usepackage{cleveref}
%\usepackage[numbers]{natbib}
\usepackage{units}
\usepackage{wasysym}
%\usepackage{caption}
%\usepackage{subcaption}

@@

\newcommand{\sam}[1]{\textcolor{red}{(Sam: #1)}}
\newcommand{\tom}[1]{\textcolor{blue}{(Tom: #1)}}
\newcommand{\mo}[1]{\textcolor{green}{(Mohamed: #1)}}

@@

As far as we know, the only published study into Regge pole spectra for relativistic stellar models was that of Chandrasekhar and Ferrari in 1991 \cite{ChandrasekharIV:1992ey}. %They used Regge pole theory to calculate the flow of gravitational energy through a star. They restricted their attention to resonant modes, i.e. poles with small imaginary parts. For the axial sector these are only present for UCOs. For the polar sector the situation is more complicated, as there are long lived fluid pulsation modes.  
They considered resonant modes (poles with small imaginary parts corresponding to trapped $w$-modes or fluid modes), and used CAM theory to calculate the flow of gravitational energy through a star. More literature exists on Regge pole theory applied to black hole scattering, for example Andersson and Thylwe have described scattering of scalar waves by a Schwarzschild BH utilising CAM techniques \cite{Andersson:1994rk}. They associate the BH Regge poles with ``surface waves'' localised near the unstable light ring at $r=3M$. Later, Decanini and collaborators showed that the Schwarzschild BH QNMs are in fact Breit-Wigner type resonances generated by said surface waves. We see here that CAM theory is a powerful approach to BH physics, that provides a clear and robust physical picture. It seems reasonable to expect that CAM may also aid in our understanding of other compact objects such as UCOs and neutron stars. With this in mind, we now turn our attention to calculating the Regge pole spectrum for compact bodies.
\tom{Some of this paragraph may be more suited to the introduction?}

 \subsection{Numerical method}\label{subsec:method}

@@

Leins Nollert and Soffel \cite{Leins:1993zz} have developed a method for calculating the polar QNM frequencies of spherically symmetric spacetimes. Benhar, Berti and Ferarri (BBF) have extended this method to the axial sector. Here we shall present a generalisation the BBF method for axial \emph{and} scalar QNM frequencies ($s=2$ and $s=0$ respectively). The method is equally valid, \textit{mutatis mutandis}, for finding Regge poles $\{\lambda_n(\omega)= \ell_n +1/2\}$ of $ S_{\ell}(\omega)$.  The solution of the Regge-Wheeler equation is written in a power-series form as follows:
%
\begin{eqnarray}\label{Power_series}
% \nonumber % Remove numbering (before each equation)
 \phi_{\omega_\ell}(r)&=&\left(\frac{r}{2M}-2M\right)^{i 2M\omega} \,e^{i\omega r} \sum_{n,=0}^{+\infty}a_n\left(1-\frac{b}{r}\right)^n \nonumber \\
                      &=& e^{i\omega r_*(r)}\sum_{n,=0}^{+\infty}a_n\left(1-\frac{b}{r}\right)^n
\end{eqnarray}
where $r = b$ is some point outside the star. By substituting \eqref{Power_series} in the Regge-Wheeler equation, it can be shown that the coefficients $a_n$ satisfy a four–term recurrence relation of the form:
\begin{equation}\label{Recurrence_4_terms}
\alpha_n a_{n+1} + \beta_n a_{n} +\gamma_{n} a_{n-1} +\delta_{n} a_{n-2}  = 0, \quad \forall n\geq 2
\end{equation}
where
\begin{eqnarray}\label{Coeffs_3_termes}
 && \alpha_n = n (n+1)\left(1-\frac{2M}{b}\right)   \\
 && \beta_n  = 2i\omega b n + 3\left(\frac{2M}{b}\right)n^2-2n^2  \\
 && \gamma_n = \left(1-\frac{6M}{b}\right)n(n-1)-\beta_s\left(\frac{2M}{b}\right)-\ell(\ell+1)  \\
 && \delta_n = \left(\frac{2M}{b}\right)\left(n(n-2)+\beta_s\right)
\end{eqnarray}
To determine the initial conditions $a_0$ and $a_1$, we impose the continuity of $\phi_{\omega_\ell}(r)$ and its derivative in $r = b$, according to Eq~\eqref{Power_series}, we obtain:
\begin{eqnarray}\label{Initial_Conds}
 && a_0 = e^{i\omega r_*(b)\phi_{\omega_\ell}(b)}  \\
 && a_1 = b e^{-i\omega r_*(b)}\left(\frac{-i\omega b}{b-2M}\phi_{\omega_\ell}(b)+ \frac{d}{dr}\phi_{\omega_\ell}(r)\Big{|}_{r=b}\right)
\end{eqnarray}
Here, to obtain $\phi_{\omega_\ell}(b)$ and its derivative in $r=b$, we need to solve numerically the Eq.~\eqref{H_Radial_equation} inside the star and to continue the solution outside, up to $r=b$, by solving the Regge-wheeler equation.

@@

We do not use the gaussian elimination step. Instead we find solutions of \eqref{Recurrence_4_terms} by adapting the \textit{Hill determinant approach} employed by Majumdar and Panchapakesan~\cite{mp} to solve the black hole recurrence formula of Leaver \cite{Leaver:1985ax}. The nontrivial solutions of \eqref{Recurrence_4_terms} exist when the Hill determinant vanishes:
\begin{equation}
\label{Determinant_Hill_4_termes}
D   =  \begin{vmatrix}
   \beta_0         &  \alpha_0    &        0             &       0                    &          0                &     \ldots          &     \ldots        &    \ldots \\
   \gamma_1        &  \beta_1      &     \alpha_1     &       0                    &          0                &     \ldots          &     \ldots        &    \ldots  \\
    \delta_2       &   \gamma_2     &    \beta_2       &   \alpha_2             &         0                &     \ldots           &    \ldots         &    \ldots  \\
    \vdots         &      \ddots     &     \ddots       &    \ddots                &       \ddots          &     \ddots           &   \ldots          &    \ldots \\
     \vdots        &     \vdots      &     \delta_{n-1}      &    \gamma_{n-1}      &     \beta_{n-1}     &      \alpha_{n-1} &    \ddots         &    \ldots  \\
     \vdots        &  \vdots        &     \vdots        &     \delta_n                &    \gamma_{n}    &     \beta_{n}        &     \alpha_{n}  &    \ddots     \\
     \vdots        &  \vdots        &     \vdots        &     \vdots                &    \ddots            &     \ddots            &     \ddots       &    \ddots
                        \end{vmatrix}
       = 0.
\end{equation}
%
Letting $D_n$ be the determinant of the $n \times n$ sub matrix of $D$ (with diagonal $\{\beta_0,\beta_1,....\beta_n\}$), by \eqref{Recurrence_4_terms}
%
\begin{equation}
D_n = \beta_n D_{n-1} - \alpha_{n-1} \gamma_n D_{n-2} + \alpha_{n-1} \alpha_{n-2} \delta_n D_{n-3}.
\end{equation}
%
Regge poles (QNM frequencies) are found by fixing $\omega$ ($\lambda$) and numerically finding roots $\lambda_n$ ($\omega_n$) of $D_n$, i.e.~ $D_n(\lambda_n,\omega)=0$ ($D_n(\lambda,\omega_n)=0$). 
%Regge poles $\lambda_n$ (QNM frequencies $\omega_n$) are roots of $D_n$

 \subsection{Numerical results: the spectrum}\label{subsec:results1}
In this section we show numerical results for the Regge pole spectrum of a compact body, in two particular cases: (i) a neutron-star-like body with tenuity $R / M = 6$, and (ii) a UCO with tenuity $R / M = 2.26$ close to the Buchdahl bound. Results for the frequencies $M \omega = 3/2$ and $M \omega = 8$ are compared, for the scalar-field ($s=0$) and axial GW ($s=2$) cases.

>>>>

%\usepackage[margin=1in]{geometry}
\usepackage{float}
%\usepackage{subfig}
\usepackage{comment}
\usepackage{array}
%\usepackage{varioref}
%\usepackage[hidelinks]{hyperref}
%\usepackage{cleveref}
%\usepackage[numbers]{natbib}
\usepackage{units}
\usepackage{wasysym}
%\usepackage{caption}
%\usepackage{subcaption}
\usepackage{xcolor}

@@

\newcommand{\sam}[1]{\textcolor{red}{(Sam: #1)}}
\newcommand{\tom}[1]{\textcolor{blue}{(Tom: #1)}}
\newcommand{\moh}[1]{\textcolor{green}{(Mohamed: #1)}}

@@

As far as we know, the only published study into Regge pole spectra for relativistic stellar models was that of Chandrasekhar and Ferrari in 1991 \cite{ChandrasekharIV:1992ey}. %They used Regge pole theory to calculate the flow of gravitational energy through a star. They restricted their attention to resonant modes, i.e. poles with small imaginary parts. For the axial sector these are only present for UCOs. For the polar sector the situation is more complicated, as there are long lived fluid pulsation modes.
They considered resonant modes (poles with small imaginary parts corresponding to trapped $w$-modes or fluid modes), and used CAM theory to calculate the flow of gravitational energy through a star. More literature exists on Regge pole theory applied to black hole scattering, for example Andersson and Thylwe have described scattering of scalar waves by a Schwarzschild BH utilising CAM techniques \cite{Andersson:1994rk}. They associate the BH Regge poles with ``surface waves'' localised near the unstable light ring at $r=3M$. Later, Decanini, Folacci and Jensen showed that the Schwarzschild BH QNMs are in fact Breit-Wigner type resonances generated by said surface waves \cite{Decanini:2002ha}. We see here that CAM theory is a powerful approach to BH physics, that provides a clear and robust physical picture. It seems reasonable to expect that CAM may also aid in our understanding of other compact objects such as UCOs and neutron stars. With this in mind, we now turn our attention to calculating the Regge pole spectrum for compact bodies.
\tom{Some of this paragraph may be more suited to the introduction?}\\
\moh{Reference Added for Decanini, Folacci and Jensen}.
 \subsection{Numerical method}\label{subsec:method}

@@

Leins, Nollert and Soffel \cite{Leins:1993zz} have developed a method for calculating the polar QNM frequencies of spherically symmetric spacetimes. Benhar, Berti and Ferarri (BBF) have extended this method to the axial sector. Here we shall present a generalisation the BBF method for axial \emph{and} scalar QNM frequencies ($s=2$ and $s=0$ respectively). The method is equally valid, \textit{mutatis mutandis}, for finding Regge poles $\{\lambda_n(\omega)= \ell_n +1/2\}$ of $ S_{\ell}(\omega)$.  The solution of the Regge-Wheeler equation is written in a power-series form as follows:
%
\begin{eqnarray}\label{Power_series}
% \nonumber % Remove numbering (before each equation)
 \phi_{\omega_\ell}(r)&=&\left(\frac{r}{2M}-1\right)^{i 2M\omega} \,e^{i\omega r} \sum_{n,=0}^{+\infty}a_n\left(1-\frac{b}{r}\right)^n \nonumber \\
                      &=& e^{i\omega r_*(r)}\sum_{n,=0}^{+\infty}a_n\left(1-\frac{b}{r}\right)^n
\end{eqnarray}
where $r = b$ is some point outside the star. By substituting \eqref{Power_series} in the Regge-Wheeler equation, it can be shown that the coefficients $a_n$ satisfy a four–term recurrence relation of the form:
\begin{equation}\label{Recurrence_4_terms}
\alpha_n a_{n+1} + \beta_n a_{n} +\gamma_{n} a_{n-1} +\delta_{n} a_{n-2}  = 0, \quad \forall n\geq 2
\end{equation}
where
\begin{eqnarray}\label{Coeffs_3_termes}
 && \alpha_n = n (n+1)\left(1-\frac{2M}{b}\right)   \\
 && \beta_n  = 2i\omega b n + 3\left(\frac{2M}{b}\right)n^2-2n^2  \\
 && \gamma_n = \left(1-\frac{6M}{b}\right)n(n-1)-\beta_s\left(\frac{2M}{b}\right)-\ell(\ell+1)  \\
 && \delta_n = \left(\frac{2M}{b}\right)\left(n(n-2)+\beta_s\right)
\end{eqnarray}
To determine the initial conditions $a_0$ and $a_1$, we impose the continuity of $\phi_{\omega_\ell}(r)$ and its derivative in $r = b$, according to Eq~\eqref{Power_series}, we obtain:
\begin{eqnarray}\label{Initial_Conds}
 && a_0 = e^{i\omega r_*(b)\phi_{\omega_\ell}(b)}  \\
 && a_1 = b e^{-i\omega r_*(b)}\left(\frac{-i\omega b}{b-2M}\phi_{\omega_\ell}(b)+ \frac{d}{dr}\phi_{\omega_\ell}(r)\Big{|}_{r=b}\right)
\end{eqnarray}
Here, to obtain $\phi_{\omega_\ell}(b)$ and its derivative in $r=b$, we need to solve numerically the Eq.~\eqref{H_Radial_equation} inside the star and to continue the solution outside, up to $r=b$, by solving the Regge-wheeler equation.

@@

We do not use the gaussian elimination step. Instead we find solutions of \eqref{Recurrence_4_terms} by adapting the \textit{Hill determinant approach} employed by Majumdar and Panchapakesan~\cite{mp} to solve the black hole recurrence formula of Leaver \cite{Leaver:1985ax}. The nontrivial solutions of \eqref{Recurrence_4_terms} exist when the Hill determinant vanishes:
\begin{equation}
\label{Determinant_Hill_4_termes}
D   =  \begin{vmatrix}
   \beta_0         &  \alpha_0    &        0             &       0                    &          0                &     \ldots          &     \ldots        &    \ldots \\
   \gamma_1        &  \beta_1      &     \alpha_1     &       0                    &          0                &     \ldots          &     \ldots        &    \ldots  \\
    \delta_2       &   \gamma_2     &    \beta_2       &   \alpha_2             &         0                &     \ldots           &    \ldots         &    \ldots  \\
    \vdots         &      \ddots     &     \ddots       &    \ddots                &       \ddots          &     \ddots           &   \ldots          &    \ldots \\
     \vdots        &     \vdots      &     \delta_{n-1}      &    \gamma_{n-1}      &     \beta_{n-1}     &      \alpha_{n-1} &    \ddots         &    \ldots  \\
     \vdots        &  \vdots        &     \vdots        &     \delta_n                &    \gamma_{n}    &     \beta_{n}        &     \alpha_{n}  &    \ddots     \\
     \vdots        &  \vdots        &     \vdots        &     \vdots                &    \ddots            &     \ddots            &     \ddots       &    \ddots
                        \end{vmatrix}
       = 0.
\end{equation}
%
Letting $D_n$ be the determinant of the $n \times n$ sub matrix of $D$ (with diagonal $\{\beta_0,\beta_1,....\beta_n\}$), by \eqref{Recurrence_4_terms}
%
\begin{equation}
\label{derecurrence_4_termes}
D_n=\beta_n D_{n-1} - \gamma_{n}\alpha_{n-1}D_{n-2} + \delta_n \alpha_{n-1} \alpha_{n-2} D_{n-3}
\end{equation}
%
with the initial conditions
 \begin{equation}
 \label{Determinant_initial_conds}
 \begin{split}
    D_0 &=\beta_0, \\
    D_1 &=\beta_1\beta_0-\gamma_1\alpha_0  \\
    D_2 &=\beta_0(\beta_1 \beta_2 - \alpha_1 \gamma_2)- \alpha_0(\alpha_1 \delta_2 -\gamma_1 \beta_2)
 \end{split}
 \end{equation}
% 
Regge poles (QNM frequencies) are found by fixing $\omega$ ($\lambda$) and numerically finding roots $\lambda_n$ ($\omega_n$) of $D_n$, i.e.~ $D_n(\lambda_n,\omega)=0$ ($D_n(\lambda,\omega_n)=0$).
%Regge poles $\lambda_n$ (QNM frequencies $\omega_n$) are roots of $D_n$
\\
\moh{initial conditions added for $D_n$}
 \subsection{Numerical results: the spectrum}\label{subsec:results1}
In this section we show numerical results for the Regge pole spectrum of a compact body, in two particular cases: (i) a neutron-star-like body with tenuity $R / M = 6$, and (ii) a UCO with tenuity $R / M = 2.26$ close to the Buchdahl bound. Results for the frequencies $M \omega = 3/2$ and $M \omega = 8$ are compared, for the scalar-field ($s=0$) and axial GW ($s=2$) cases.
